---
title: "Getting Started with hvtiRimputation"
author: "John Ehrlinger"
date: today
format:
  html:
    toc: true
vignette: >
  %\VignetteIndexEntry{Getting Started with hvtiRimputation}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Why keep a record of what you filled?

Most of us have filled a missing covariate with its mean and moved on. In SAS
that was `PROC STANDARD ... REPLACE`, usually through the `imputsub` macro, and
the job ended with a complete dataset and no trace of which values had been
missing. Months later a reviewer asks how many patients were in the model only
because a covariate was filled in, or whether one particular patient's value
was observed, and the dataset can't tell you.

`hvtiRimputation` fills missing values by a stated method and hands back the
filled data together with a record of exactly what it changed. The record is a
logical matrix with one row per row of your data and one column per imputed
variable, `TRUE` where a value was filled. Every count, indicator and
attrition column in the package is read from that matrix.

The scope is deliberately small. A data frame goes in, and a data frame and a
record come out. The package knows nothing about the warehouse, the build or
any study, so the study context (manifests, cohort definitions, provenance of
the input) stays with you.

This vignette is for the CORR biostatistician who already knows R and has
used the SAS macros. It walks one small synthetic example from the fill to the
attrition columns a CONSORT diagram wants, then shows the multiple-imputation
prototype.

## Installation

```{r install, eval = FALSE}
# install.packages("pak")
pak::pak("ehrlinger/hvtiRimputation")
```

```{r library}
library(hvtiRimputation)
```

## A small example

Eight synthetic rows: three covariates with gaps, and an outcome that is
missing for one row. Nothing here comes from a real patient.

```{r data}
dat <- data.frame(
  age     = c(64, 71, NA, 58, 66, NA, 73, 69),
  bmi     = c(27.1, NA, 31.4, 24.8, NA, 29.0, 26.3, 30.2),
  ef      = c(55, 60, 45, NA, 50, 58, 62, 40),
  outcome = c(0, 1, 0, 0, 1, NA, 1, 0)
)
dat
```

Only three rows are complete. Listwise deletion would keep rows 1, 7 and 8.

## Fill, then read the record

`impute_mean()` replaces each missing value in `vars` with that variable's
mean over its observed values. Observed values are never touched.

```{r impute-mean}
out <- impute_mean(dat, vars = c("age", "bmi", "ef"))
out
```

`vars` has no default. `PROC STANDARD` with no `VAR` statement processes every
numeric column, and we chose not to inherit that: an imputation that picks its
own variable list is the mistake the record exists to catch. Here `outcome` is
numeric and is left alone because we did not name it.

The completed data, and the matrix that says which cells in it were filled:

```{r read-record}
imputed_data(out)
imputed_matrix(out)
```

Row 3's `age` of 66.8 is the mean of the six observed ages. The matrix is what
lets you tell it apart from an observed 66.8, and it answers the audit
question directly: was this patient's value imputed? A count can't answer that,
though the matrix reduces to any count you need. `summary()` is one such view:

```{r summary}
summary(out)
```

## Which rows did imputation keep?

Mean imputation is usually standing in for listwise deletion, so the question
a CONSORT diagram asks is how many rows entered the analysis only because
something was filled. Three row-level columns answer it.

```{r row-columns}
data.frame(
  imputed_any        = imputed_any(out),
  complete_case_pass = complete_case_pass(out),
  analysis_pass      = analysis_pass(out),
  kept_by_imputation = kept_by_imputation(out)
)
```

`complete_case_pass()` is read from the data *before* anything was filled, over
`cc_vars` (every column, by default). `analysis_pass()` is the same test after
the fill. `kept_by_imputation()` is the row that fails the first and passes the
second.

Look at row 6. Its `age` was filled, so `imputed_any()` is `TRUE`, but its
`outcome` is still missing, so it never enters the analysis. That is why the
tempting shortcut gives the wrong count:

```{r overcount}
sum(imputed_any(out) & !complete_case_pass(out)) # overcounts row 6
sum(kept_by_imputation(out))                     # the supported answer
```

The two agree only when imputation completes every row. Use
`kept_by_imputation()`. It is also the number worth checking a port against:
a port that mean-imputes the wrong variable list can still reach the right
final row count, but it won't reach the right `kept_by_imputation()` total.

## Indicators for a model

To carry "this value was missing" into a model, generate indicator columns
from the record. The prefix is your choice; `"ms_"` reproduces the SAS naming
when a study used it.

```{r indicators}
imputation_indicators(out)
cbind(imputed_data(out), imputation_indicators(out, prefix = "ms_"))
```

The indicators are generated from the matrix, never stored separately, so they
can't drift from it.

## How it was produced

`imputation_provenance()` records the method, `m`, the variables, the fill
values and the package version. Save it alongside the analysis.

```{r provenance}
prov <- imputation_provenance(out)
prov[c("method", "m", "vars", "cc_vars", "fill_values", "version")]
```

## What it refuses

The record must never claim a fill that did not happen, so the package refuses
inputs a more relaxed function would quietly accept. A variable missing in
every row has no mean to fill from:

```{r refuse}
all_missing <- transform(dat, ef = NA_real_)
tryCatch(
  impute_mean(all_missing, vars = c("age", "ef")),
  error = conditionMessage
)
```

The same goes for a non-numeric variable, a name that is not in the data, a
column whose mean is not finite, a classed numeric such as `integer64` or
`Date`, duplicated column names and a zero-row frame. Each is an error, not a
warning. A refusal costs you a line of code; a wrong record costs you an audit
nobody can pass, because nothing downstream can detect it.

## Multiple imputation (prototype)

`impute_multiple()` is the R form of `PROC MI`, which the `mult_imput` macro
wraps. It draws `m` completed datasets with `mice` and returns the same shape
of record. It is a prototype: designed and tested, not yet hardened or
released, and it does **not** pool. Fitting a model to each draw and combining
the results by Rubin's rules is still your job.

It is a separate function from `impute_mean()` on purpose. The two are
different methods with different inferential properties, and there is no
single `impute()` with a `method=` argument that could quietly run one when
you expected the other.

```{r impute-multiple}
mi <- impute_multiple(
  dat,
  vars  = c("age", "bmi", "ef"),
  m     = 2,
  maxit = 2,
  seed  = 20261007
)
mi
```

`m` and `seed` are both required. Copies of the SAS macros declare different
`NIMPUTE` defaults, and they don't even agree on whether the default is 1, so
there is no default to inherit. An unseeded draw can't be reproduced. You can set the
seed once per session with `options(hvtiRimputation.seed = )` instead of
passing it every call.

`imputed_data()` needs to be told which draw you want, either one completed
dataset or all of them stacked long with `.imp` and `.id` columns:

```{r multi-data}
imputed_data(mi, imputation = 1)
head(imputed_data(mi, imputation = "long"))
```

There is no default for `imputation`, because averaging the draws cell by cell
is not a valid way to combine them. The record and the row-level columns read
the same as before, since which cells were missing does not change from draw
to draw:

```{r multi-record}
imputed_matrix(mi)
sum(kept_by_imputation(mi))
```

## Where to go next

- `?impute_mean` and `?impute_multiple`, each with a **Divergences from SAS**
  section listing where the R port differs from the macro and why.
- `?hvti_imputation` and `?hvti_imputation_multi` for the full shape of the
  returned objects.
- The [package README](https://github.com/ehrlinger/hvtiRimputation#readme)
  for what the package promises about reproducing SAS. In short: the
  arithmetic is matched, the defaults deliberately are not, and a study being
  reproduced takes its `m` from that study's own saved output.
